---
title: "MCS 8920 — Environment Smoke Test"
author: "Minhajur Rahman"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    fig-width: 7
    fig-height: 3.5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
# If any of these fail to load, that package didn't install --
# note which one and stop here rather than debugging downstream errors.
library(tidyverse)
library(haven)
library(survey)
library(srvyr)
library(gtsummary)
library(janitor)
library(here)
library(naniar)
library(patchwork)
library(sessioninfo)
```

# What this document tests

Each section below exercises one part of the toolchain. If the whole
thing renders to PDF, the environment is ready for the real project.

| Section | What it proves |
|---|---|
| 2 | R runs, `here()` resolves the project root |
| 3 | `haven` reads and writes NHANES-style `.XPT` files |
| 4 | `survey` + `srvyr` handle a complex design correctly |
| 5 | `gtsummary` produces a table that survives LaTeX |
| 6 | `ggplot2` + `patchwork` render figures into the PDF |
| 7 | `naniar` works (missingness is a graded decision later) |
| 8 | Session is reproducible and reportable |

The fact that you are reading this as a PDF at all proves Quarto and
TinyTeX are working.

# Project root and paths

`here()` anchors paths to the `.Rproj` file rather than to the working
directory. This matters because a Quarto chunk's working directory is
not always what you expect, and hardcoded paths are the most common
reason someone else's repo fails to render on your machine.

```{r}
#| label: paths
here()
file.exists(here("data", "raw"))
```

# haven: reading and writing XPT

NHANES ships as SAS transport files. Rather than depend on the CDC
being reachable right now, this round-trips a small file: write it,
read it back, confirm it survived. Same code path `read_xpt()` will
take on the real data.

```{r}
#| label: xpt-roundtrip
demo <- tibble(
  seqn = 1:6,
  age  = c(34, 51, 22, 67, 45, 29),
  sex  = c(1, 2, 2, 1, 2, 1)
)

xpt_path <- here("data", "raw", "demo_test.xpt")
write_xpt(demo, xpt_path)

back <- read_xpt(xpt_path)
back

identical(dim(demo), dim(back))
```

`R/00_download_data.R` attempts a real CDC download separately — run it
if you also want to confirm network access, but it is not needed for
this render.

# Complex survey design

This uses the `api` data bundled with the `survey` package: a real
cluster sample with weights (`pw`), a clustering variable (`dnum`), and
a finite population correction (`fpc`). Structurally the same problem as
NHANES's `SDMVSTRA` / `SDMVPSU` / `WTMECPRP`, without needing a download.

```{r}
#| label: survey-design
data(api)

des <- apiclus1 |>
  as_survey_design(ids = dnum, weights = pw, fpc = fpc)

des |>
  summarise(
    mean_api00 = survey_mean(api00, vartype = "ci"),
    n_obs      = unweighted(n())
  )
```

The point of the comparison below: the naive mean ignores the design and
gives a different answer with a much narrower, and wrong, interval.

```{r}
#| label: naive-vs-weighted
tibble(
  estimate = c("Naive (unweighted)", "Design-based"),
  api00    = c(mean(apiclus1$api00), coef(svymean(~api00, des)))
)
```

```{r}
#| label: janitor-demo
apiclus1 |> tabyl(stype)
```

# Table output

`gtsummary` renders through `gt` by default, which is fussy in LaTeX.
`as_kable()` is the reliable path for PDF output — worth knowing before
the real report is due.

```{r}
#| label: table
svy <- svydesign(id = ~dnum, weights = ~pw, fpc = ~fpc, data = apiclus1)

tbl_svysummary(
  svy,
  by      = stype,
  include = c(api00, ell, meals)
) |>
  as_kable()
```

# Figures

```{r}
#| label: figures
#| fig-cap: "Left: school-level poverty proxy against test score. Right: score by school type."
p1 <- ggplot(apiclus1, aes(x = meals, y = api00)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "% eligible for subsidised meals", y = "API score") +
  theme_minimal()

p2 <- ggplot(apiclus1, aes(x = stype, y = api00)) +
  geom_boxplot() +
  labs(x = "School type", y = NULL) +
  theme_minimal()

p1 + p2
```

# Missingness

How missing data is handled is one of the decisions that has to be
defended in `DECISIONS.md`, so confirm the tooling works now.

```{r}
#| label: missingness
apiclus1 |>
  select(api00, ell, meals, avg.ed, hsg) |>
  gg_miss_var() +
  theme_minimal()
```

# Session information

Paste this into an issue if you and a teammate get different results.

```{r}
#| label: sessioninfo
session_info()
```
